\documentclass[12pt, letterpaper]{article}

\usepackage{spreamble}

\title{CDA3101}
\author{Ian Zou}

\begin{document}

\maketitle

\part*{Von Neumann Architecture}

Instructions and data are stored in the same memory space. Single-memory, single-bus architecture.

\subsection*{Von Neumann Bottleneck}

Tasks are completed sequentially due to the single bus, and thus the speed of this single bus is a performance bottleneck.

\begin{itemize}
    \item \textbf{CU (Control Unit)}: Manages how the processor works by sending control signals, controlling input/output operations, and fetching instructions for execution.
    \item \textbf{ALU (Arithmetic Logic Unit)}: Performs arithmetic and logic calculations.
    \item \textbf{Memory Unit}: Consists of the main memory (RAM) and the registers. Single-memory system that holds both data and instructions.
    \item \textbf{Registers}: High-speed temporary storage on the CPU.
\end{itemize}

\part*{Measuring Performance}

\section*{Orders of Magnitude}
\begin{table}[h]
    \begin{center}
        \begin{tabular}{|p{2cm}|p{3cm}|}
            \hline
            \textbf{nano} & $ 10^{-9} $ \\
            \hline
            \textbf{micro} & $ 10^{-6} $ \\
            \hline
            \textbf{milli} & $ 10^{-3} $ \\
            \hline
            \textbf{kilo} & $ 10^3 $ \\
            \hline
            \textbf{mega} & $ 10^6 $ \\
            \hline
            \textbf{giga} & $ 10^9 $ \\
            \hline
            \textbf{tera} & $ 10^{12} $ \\
            \hline
        \end{tabular}
    \end{center}
\end{table}
\FloatBarrier

\section*{Units of Measure}
\begin{itemize}
    \item \textbf{Frequency}: \textit{Hertz}, or cycles per second. Used for \textbf{clock rate}.
    \item \textbf{Period}: Seconds per cycle. Used for \textbf{clock cycle}.
    \item \textbf{CPI}: Clocks per instruction. Highlights a CPU's \textbf{efficiency}.
\end{itemize}

Use \textit{dimensional analysis} to get to the desired unit.

\section*{Speedup}

$ \textrm{speedup} = \frac{t_{\textrm{old}}}{t_{\textrm{new}}} $

Where $ t_{\textrm{old}} $ and $ t_{\textrm{new}} $ are the times the program took before and after the upgrade, \textit{in seconds}.

\part*{discrete shit}

\section*{Two's Complement}

\textbf{Two's complement} is the standard for representing signed integers.

\begin{itemize}
\item \textbf{Most significant bit (MSB)} used as a sign bit.
\item For an $n$-bit word, the range of representable numbers is $ -2^{n-1} $ to $ 2^{n-1} - 1 $.
\end{itemize}

\subsection*{Two's Complement Representation}

\begin{enumerate}
    \item Take the binary of the absolute value of the integer.
    \item Flip the bits \textit{(one's complement)}.
    \item Add \texttt{1}.
\end{enumerate}

\part*{Assembly Structures}

\begin{itemize}
    \item Local variables stored on \textbf{stack} (either static or dynamic).
    \item Instructions and programs are stored in \textbf{text}.
\end{itemize}

\part*{Byte Order}

\begin{itemize}
    \item \textbf{Big-Endian (BE)}: Stores most significant byte first.
    \item \textbf{Little-Endian (LE)}: Stores most significant byte last.
\end{itemize}

\part*{RISC-V Instructions}

\begin{table}[h]
\begin{center}
    \begin{tabular}{|p{4cm}|p{6cm}|p{5cm}| }
        \hline
        \textbf{Mnemonic} & \textbf{Instruction} & \textbf{Description} \\
        \hline
        \texttt{add rd, rs1, rs2 } & Add & \texttt{rd <- rs1 + rs2}\\
        \hline
        \texttt{sub rd, rs1, rs2 } & Subtract & \texttt{rd <- rs1 - rs2}\\
        \hline
        \texttt{addi rd, rs1, imm12 } & Add (immediate r-value) & \texttt{rd <- rs1 + imm12}\\
        \hline
        \texttt{add rd, rs1, rs2 } & Add & \texttt{rd <- rs1 + rs2}\\
        \hline
    \end{tabular}
\end{center}
\end{table}

\FloatBarrier

\part*{RISC-V Machine Code}

\texttt{32-bit} instructions split into fields.

\section*{R-Format Instruction Structure}

\begin{itemize}
    \item \texttt{func7} (\texttt{7 bits})
        \begin{itemize}
            \item Basic operation of the instruction.
        \end{itemize}
    \item \texttt{rs2} (\texttt{5 bits})
        \begin{itemize}
            \item Second source register.
        \end{itemize}
    \item \texttt{rs1} (\texttt{5 bits})
        \begin{itemize}
            \item First source register.
        \end{itemize}
    \item \texttt{func3} (\texttt{3 bits})
        \begin{itemize}
            \item Basic operation of the instruction.
        \end{itemize}
    \item \texttt{rd} (\texttt{5 bits})
        \begin{itemize}
            \item Destination register.
        \end{itemize}
    \item \texttt{opcode} (\texttt{7 bits})
        \begin{itemize}
            \item Basic operation of the instruction.
        \end{itemize}
\end{itemize}

\section*{I-Format Instruction Structure}

\begin{itemize}
    \item \texttt{imm} (\texttt{12 bits})
        \begin{itemize}
            \item Immediate value.
        \end{itemize}
    \item \texttt{rs1} (\texttt{5 bits})
        \begin{itemize}
            \item First source register.
        \end{itemize}
    \item \texttt{func3} (\texttt{3 bits})
        \begin{itemize}
            \item Basic operation of the instruction.
        \end{itemize}
    \item \texttt{rd} (\texttt{5 bits})
        \begin{itemize}
            \item Destination register.
        \end{itemize}
    \item \texttt{opcode} (\texttt{7 bits})
        \begin{itemize}
            \item Basic operation of the instruction.
        \end{itemize}
\end{itemize}

\section*{S-Format Instruction Structure}

Same as \textit{R-Format Instruction Structure}, but \texttt{func7} is replaced with immediate-value address to some memory.

\part*{Building a Datapath}

\section*{Digital Logic}

Representation of signals and states of digital circuit with \textbf{binary}.

\medskip
{\centering

$ z = f(x, y), $ where $ x, y, z \in \begin{Bmatrix} 0, 1 \end{Bmatrix} $

}
\medskip

\subsection*{Logic Gates}

\begin{itemize}
    \item \textbf{NOT Gate}: $ Z = X' $, $ Z = \overline{X} $
    \item \textbf{AND Gate}: $ Z = X \times Y $
    \item \textbf{OR Gate}: $ Z = X' + Y $
    \item \textbf{NAND Gate} % TODO
    \item \textbf{NOR Gate}
    \item \textbf{XOR Gate}
\end{itemize}

\subsection*{Building a Circuit}

% TODO

\subsection*{Comparator Circuit}

\textbf{Comparator circuits} compare bits of binary numbers to see if they are equal.

\subsection*{Multiplexer Circuit}

\textbf{Multiplexer circuits} select one data channel from many input channels.

\begin{itemize}
    \item Maps the selected channel to the output channel.
    \item Control signals used to select channel.
    \item Can represent $ 2^n $ channels, where $ n $ is the number of control bits.
\end{itemize}

\subsection*{Decoder Circuit}

\textbf{Decoder circuits} have many possible outputs that can be activated.

\begin{itemize}
    \item \textbf{Encoder} chooses one of the outputs.
    \item Control signals used to make the choice.
\end{itemize}

\subsection*{Sum of Products}

Any logic function can be described as \texttt{OR}-ing multiple \texttt{AND} results.
\medskip

\medskip
{\centering

$ Z = (X \textrm{AND} Y) \textrm{OR} (X' \textrm{AND} Y') $

}
\medskip

% TODO: full instructions

\subsection*{Product of Sums}

Any logic function can be described as \texttt{AND}-ing multiple \textrm{OR} results.

\medskip
{\centering

$ Z = (X \textrm{OR} Y) \textrm{AND} (X' \textrm{OR} Y') $

}
\medskip

% TODO: full instructions

\part*{Single-Cycle Datapath}

\section*{Datapath Design}

The \textbf{Instruction Set Architecture} determines many aspects of implementation, and these strategies affect \textbf{clock rate} and \textbf{CPI}.

\section*{Logic Design}

\begin{itemize}
    \item \textbf{Combinational Elements}
        \begin{itemize}
            \item Combination of gates, ALU.
            \item Outputs depend only on inputs.
        \end{itemize}
    \item \textbf{State Elements}
        \begin{itemize}
            \item Memory, storage.
            \item Data read from register is state value at previous clock cycle.
        \end{itemize}
    \item \textbf{Datapath}
        \begin{itemize}
            \item Reads an instruction.
            \item Executes instructions.
            \item Writes registers.
            \item \texttt{PC} stores address of current instruction.
            \item \textbf{Instruction Memory} stores the instruction.
        \end{itemize}
\end{itemize}

\section*{Register File}

\begin{itemize}
    \item Register numbers are taken from the instruction and sent to the register file.
    \item Data to be provided is sent through the \textbf{read ports}.
    \item Signal with the register number to be written, and a \textbf{write port} through which the data will be sent.
    \item \textbf{Read Data}: State of register in the previous clock cycle.
    \item \textbf{Write Data}: Changed for the next clock cycle.
\end{itemize}

\section*{Datapath Elements for R-Format}

\begin{enumerate}
    \item Instruction is read from \textbf{instruction memory}.
    \item Instruction is split by Decoder-Logic into \texttt{opcode}, \texttt{rs1}, \texttt{rs2}, and \texttt{rd} fields.
    \item Register contents for \texttt{rs1} and \texttt{rs2} are read from \textbf{Register File}.
    \item \textbf{ALU} computes using \texttt{rs1} and \texttt{rs2} registers and produces result.
    \item Result is written to \textbf{Register File} at \texttt{WriteData} port.
\end{enumerate}

\section*{Datapath Elements for S-Format}

\begin{enumerate}
    \item Instruction is read from \textbf{instruction memory}.
    \item Instruction is split by Decoder-Logic into \texttt{opcode}, \texttt{src} or \texttt{dst}, \texttt{offset}, and \texttt{base}.
    \item \texttt{offset} sent to \textbf{Sign Extension}.
    \item \textit{Load} output sent to \texttt{ReadData1}. \textit{Store} source is sent to \texttt{ReadData2}.
    \item \textbf{ALU} computes memory address.
    \item \textbf{ALU} output applied to data memory.
    \item \textbf{LW Only:} \textbf{Data Memory} output written to \textbf{Register File}.
\end{enumerate}

\section*{Datapath Elements for Branch}

\begin{enumerate}
    \item Instruction is read from \textbf{Instruction Memory}.
    \item Instruction is split by Decoder-Logic into \texttt{opcode}, \texttt{rs1}, \texttt{rs2}, and \texttt{imm} fields.
    \item Register contents read from \textbf{Register File} (output at \texttt{ReadData1} and \texttt{ReadData2}).
    \item \textbf{ALU} adds \texttt{PC} to sign-extended \texttt{imm} to produce \texttt{BTA}.
    \item \textbf{ALU} compares \texttt{rs1} and \texttt{rs2} registers and calculates zero output (valued 1 or 0).
\end{enumerate}

\part*{Pipelining}

Separate resource for each stage in the datapath in order to reduce the time taken for multiple consecutive instructions.

\begin{itemize}
    \item For a pipeline, each \textbf{stage} is a \textbf{cycle}.
    \item \textbf{Cycle Time}: Time of the longest stage.
    \item Each instruction takes $ \textrm{stage} \times \textrm{cycle time} $ to execute.
\end{itemize}

Pipelining enables \textbf{parallel execution} through overlapping execution of different instruction stages concurrently.

\section*{Classic Pipeline}

\begin{enumerate}
    \item \textbf{Instruction Fetch}
    \item \textbf{Register Read}
    \item \textbf{ALU Operation}
    \item \textbf{Data Access}
    \item \textbf{Register Write}
\end{enumerate}

This pipeline breaks each instruction into 5 stages. By making the \textbf{cycle time} shorter (duration of the longest stage), we can make each instruction take 5 cycles to execute.

\part*{Cache Memory}

\section*{Memory Hierarchy}

\subsection*{Maximize Memory Performance}

\begin{itemize}
    \item \textbf{Maximize} data transfers in \textbf{fast memory} (registers <-> cache)
    \item \textbf{Minimize} data transfers in \textbf{slow memory} (registers <-> main memory)
\end{itemize}

\section*{Locality}

\textbf{Locality} can be exploited to improve performance.

\begin{itemize}
    \item \textbf{Spatial Locality}: Instructions tend to be accessed sequentially - they are near each other in memory.
    \item \textbf{Temporal Locality}: Reuse of data or instructions previously accessed.
\end{itemize}

\section*{Cache Memory}

\subsection*{Primary Cache}

\textbf{L1} or \textbf{primary cache} is on the same chip as the CPU.

\begin{itemize}
    \item \textbf{Fastest memory} outside of registers.
    \item Limited in size.
\end{itemize}

\subsection*{Secondary Cache}

\textbf{L2} or \textbf{secondary cache} can be on the same chip or separate.

\begin{itemize}
    \item Larger than \textbf{L1}.
    \item Slower access time than \textbf{L1}.
\end{itemize}

\subsection*{Harvard Cache}

\textbf{Unified} or \textbf{integrated} caches contain both instructions and data.
\bigskip

\textbf{Harvard cache} is a split cache system.

\begin{itemize}
    \item \textbf{I-cache} for caching instructions.
    \item \textbf{D-cache} for caching data.
    \item Provides better \textbf{locality} at the cost of complexity.
    \item Common setups include a \textit{split} \textbf{L1} and a \textit{unified} \textbf{L2}.
\end{itemize}

\section*{Cache Organization}

Main memory is broken into \textbf{blocks}, whereas cache memory is broken into \textbf{cache lines}.

\subsection*{Blocks}

\begin{itemize}
    \item \textbf{Block Number}: $\displaystyle \frac{\textrm{address}}{\textrm{block size}} $
    \item \textbf{Offset Bits}: $ \log_2 \textrm{block size} $
\end{itemize}

\subsection*{Cache Lines}

\begin{itemize}
    \item \textbf{Cache lines} are the same size as a \textbf{memory block}.
    \item Mapping determines which line contains which block.
    \item \textbf{Number of blocks}: $\displaystyle \frac{\textrm{main memory size}}{\textrm{block size}} $
    \item \textbf{Number of lines}: $\displaystyle \frac{\textrm{cache size}}{\textrm{block size}} $
\end{itemize}

\subsection*{Associativity}

Cache is organized into some number of sets.

\begin{itemize}
    \item \textbf{Associativity}: Number of blocks per set.
    \item \textbf{N-Way Associative}: \texttt{N} blocks per set.
\end{itemize}

\subsubsection*{Fully Associative}

\begin{itemize}
    \item Any memory block can go into any empty cache line.
    \item \textbf{Address}
        \begin{itemize}
            \item Based on dividing the address into two fields: an \textbf{tag field} and \textbf{offset field}.
            \item \textbf{Offset field} width: $ \log_2 \textrm{line\_size} $
            \item \textbf{Offset} is used to specify byte or word in the \textbf{cache line}.
        \end{itemize}
    \item \textbf{Cache Hit Detection}
        \begin{itemize}
            \item Check all \textbf{cache lines} in parallel.
            \item If a stored tag matches the \textbf{tag field}, we detect a \textbf{cache hit}.
        \end{itemize}
    \item \textbf{Block Loading}
        \begin{itemize}
            \item Block is loaded into \textbf{vacant cache line} if miss occurs.
            \item If no line is vacant, use a \textbf{replacement policy} to evict a line. The most common is \textbf{Least Recently Used (LRU)}.
        \end{itemize}
    \item Requires hardware to compare all lines in parallel.
\end{itemize}

\subsubsection*{Direct Mapped}

\begin{itemize}
    \item Each block has a direct mapping to a cache line.
    \item \textbf{Mapping}
        \begin{itemize}
            \item Maps \textbf{block number} $ B $ to line $ N $.
            \item $ L = B \: \textrm{mod} \: N $.
            \item Multiple \textbf{blocks} may map to the same \textbf{cache line}.
        \end{itemize}
    \item \textbf{Address}
        \begin{itemize}
            \item \textbf{Offset Field}: Specify byte or word in the \textbf{cache line}.
                \begin{itemize}
                    \item $ \log_2 \textrm{line\_size} $
                \end{itemize}
            \item \textbf{Line Field}: Selects a unique line of cache.
                \begin{itemize}
                    \item $ \log_2 \textrm{lines} $
                \end{itemize}
            \item \textbf{Tag Field}: Used to identify currently loaded \textbf{block}.
                \begin{itemize}
                    \item $ \texttt{bits} - \textrm{LW} - \textrm{OW} $
                \end{itemize}
        \end{itemize}
    \item \textbf{Cache Hit Detection}
        \begin{itemize}
            \item Visit the \textbf{cache line} as specified by the \textbf{line field}.
            \item If the stored tag matches the \textbf{tag field}, we detect a \textbf{cache hit}.
        \end{itemize}
    \item \textbf{Block Loading}
        \begin{itemize}
            \item On \textbf{cache miss}, load the \textbf{cache line} with the \textbf{block}.
            \item Prone to unused \textbf{vacant lines} if accessed \textbf{blocks} map to the same \textbf{line}.
            \item If no line is vacant, use a \textbf{replacement policy} to evict a line.
        \end{itemize}
\end{itemize}

\subsection*{Set Associative}

\begin{itemize}
    \item Divides \textbf{cache lines} into several \textbf{sets}.
    \item \textbf{Address}
        \begin{itemize}
            \item \textbf{Tag Field}:  Used to identify the \textbf{line} within the \textbf{set} that contains the \textbf{block}.
                \begin{itemize}
                    \item $ \texttt{bits} - \texttt{SW} - \texttt{OW} $
                \end{itemize}
            \item \textbf{Set Field}: Selects the \textbf{set} of blocks.
                \begin{itemize}
                    \item $ \log_2 (sets) $
                \end{itemize}
            \item \textbf{Offset Field}: Specify byte or word in the \textbf{cache line}.
                \begin{itemize}
                    \item $ \log_2 (block\_size) $
                \end{itemize}
        \end{itemize}
    \item \textbf{Cache Hit Detection}
        \begin{itemize}
            \item Visit the \textbf{set} as specified by the \textbf{set field}.
            \item Check all \textbf{cache lines} in the \textbf{set} in parallel.
            \item If a stored tag matches the \textbf{tag field}, we detect a \textbf{cache hit}.
        \end{itemize}
    \item \textbf{Block Loading}
        \begin{itemize}
            \item \textbf{Block} can be loaded in any \textbf{vacant cache line} in specified \textbf{set}.
        \end{itemize}
\end{itemize}

\begin{itemize}
    \item 
\end{itemize}

\section*{Cache Read}

\begin{itemize}
    \item \textbf{Cache Hit}: Data is found at a given memory level.
        \begin{itemize}
            \item \textbf{Hit Rate}: Percent of references found at a given memory level.
            \item \textbf{Hit Ratio}: $ \displaystyle \frac{\textrm{hits}}{\textrm{total\_refs}} $
        \end{itemize}
    \item \textbf{Cache Miss}: Data is not found.
        \begin{itemize}
            \item \textbf{Miss Rate}: $ 100 - \textrm{hit\_rate} $
            \item \textbf{Miss Ratio}: $ 1 - \textrm{hit\_ratio} $
            \item \textbf{Miss Penalty}: Time required to process a miss.
        \end{itemize}
\end{itemize}

\subsection*{Sources of Cache Misses}

\begin{itemize}
    \item \textbf{Compulsory Misses}: Initially empty cache.
    \item \textbf{Capacity}: Limited size of cache.
    \item \textbf{Conflict}: Different blocks mapping to the same cache line.
\end{itemize}

\subsection*{Lookaside Cache}

\begin{itemize}
    \item Accesses to cache and to main memory occur in parallel.
    \item Main memory access is cancelled if \textbf{cache hit} occurs.
    \item Lowers average memory access time at the cost of increase CPU to memory traffic.
\end{itemize}

\subsection*{Look Through Cache}

\begin{itemize}
    \item \textbf{Primary cache} checked first. \textbf{Secondary cache} checked if \textbf{cache miss} on \textbf{L1}.
    \item Only accesses main memory if \textbf{cache miss} on all cache memory.
    \item Avoids unneeded CPU-to-memory traffic, but increases average memory access time.
\end{itemize}

\section*{Cache Writing}

\begin{itemize}
    \item \textbf{Write Through}: Memory updated in parallel with every write. \textcolor{red}{Defeats the purpose of having a cache.}
    \item \textbf{Write Back}: Updates only the cache copy of the item.
        \begin{itemize}
            \item Main memory is updated when the item is removed from the cache.
            \item Must record whether cache line was written (dirty).
            \item May cause issues with concurrent access (shared memory multiprocessors).
        \end{itemize}
\end{itemize}

\end{document}
