\documentclass[12pt, letterpaper]{article}
\usepackage{../spreamble}
\usepackage{tabularx}

\title{MAS3114}
\author{Ian Zou}

\begin{document}

\maketitle

\part*{Unit 1 - Linear Systems \& Matrices}

\section*{Lecture 1}

\subsection*{Linear Equations}

A \textbf{linear equation} is in the form:

\medskip
{\centering

$ a_1 x_1 + a_2 x_2 + . . . + a_n x_n = b_1 $

}
\medskip

Where $ x_1, x_2, \hdots, x_n $ are variables, $ a_1, a_2, ..., a_n $ and $ b $ are scalars, and $ n $ is a positive integer.

\medskip

Examples of linear equations include:
\begin{itemize}
    \item $ 5x_1 + 4x_2 = 0 $
    \item $ \sqrt{5}x_1 + 4x_2 = 0 $
\end{itemize}

The following are \textit{NOT} linear equations:
\begin{itemize}
    \item $ 5\sqrt{x_1} + 4x_2 = 0 $
    \item $ 5\sqrt{x_1} + 4x_2 = 0 $
\end{itemize}

\subsection*{Linear Systems}

A \textbf{system of linear equations} (\textbf{linear system}) is a collection of one or more linear equations.
\medskip

The following is a linear system:

\medskip
{\centering

$ 2x_1 + x_2 + x_3 = 1 $
$ x_1 - 2x_2 - x_3 = 2 $

}
\medskip

A linear system is:

\begin{itemize}
    \item \textbf{Consistent}: has 1 solution or \textit{infinitely many} solutions.
    \item \textbf{Inconsistent}: has \textit{no} solutions.
\end{itemize}

Two linear systems are \textbf{equivalent} if they have the same solution set.
\medskip

\subsection*{Matrix Notation}

Used to represent a linear system.
\medskip

The following is an example of a $ 3 \times 3 $ \textbf{coefficient matrix}:

\medskip
{\centering

$ A = \begin{bmatrix}
    a_{11} & a_{12} & a_{13} \\
    a_{21} & a_{22} & a_{23} \\
    a_{31} & a_{32} & a_{33}
\end{bmatrix}  $

}
\medskip

The \textbf{size} of a matrix is the number of rows and columns. Any $ m \times n $ matrix is a 2-D array of numbers.

\begin{enumerate}
    \item \textit{Example}: Solve the system.

        {\centering

            $ x_1 - x_2 = -4 $

            $ x_1 + 2x_2 = 2 $

        }

        \bigskip

        \textit{Naive Solution}: We get the singular equation by subtracting the equations: % TODO: clarify 

        {\centering

            $ 3x_2 = 6 $

            $ x_2 = 2 $

        }

        \medskip

        We can plug it back into one of the equations to solve for $ x_1 $:

        {\centering

            $ x_1 + 2(2) = 2 $

            $ x_1 = -2 $

        }

        \medskip

        Thus, the solution set is:

        {\centering

            $\begin{bmatrix}
                -2 \\
                 2 \\
            \end{bmatrix}$

        }

\end{enumerate}

\subsection*{Elementary Row Operations}

\begin{enumerate}
    \item \textbf{Interchange}: Interchange two rows.
    \item \textbf{Scaling}: Multiply every entry in a row by a nonzero constant.
    \item \textbf{Replacement/Row Addition}: Add a multiple of one row to another row.
\end{enumerate}

Two matrices are \textbf{row equivalent} if there is a sequence of elementary row operations that transform one matrix into the other.

\textit{Row equivalent} augmented matrices have the \textit{same solution set}.
\medskip

\subsection*{Echelon Forms}

\textbf{Row Echelon Forms} help us determine whether a matrix is \textit{consistent}.

\begin{itemize}
    \item \textbf{Row Echelon Form (REF)}: Leading entries of each row are \textbf{pivots}, and these pivots propagate down and rightwards.
    \item \textbf{Reduced Row Echelon Form (RREF)}: In addition to REF, each pivot must be $ 1 $ and each pivot column should have no other nonzero entries.
\end{itemize}

If $ A $ is a matrix in REF or RREF, the first nonzero (leading) entry in each row is a pivot.
\begin{itemize}
    \item A \textbf{pivot position} in $ A $ is a location that corresponds to a pivot.
    \item A \textbf{pivot column} is a column in $ A $ that contains a pivot position.
    \item \textbf{Rank}: The number of pivot positions in the matrix $ A $.
    \item \textbf{Pivot Variables}: Variables that correspond to the pivot columns. Variables that do not correspond are called \textbf{free variables}.
    \item If the system is inconsistent, the solution set is empty even though the matrix has free variables. % TODO: fact check?
\end{itemize}

\section*{Lecture 2}

\subsection*{Row Reduction Algorithm}

\begin{enumerate}
    \item From the left, find the first pivot column.
    \item Interchange rows so that the top entry (pivot) in the column is nonzero.

        NOTE: If $ 1 $ or $ -1 $ is in the column, we prefer these entries as the pivot.
    \item Use row replacement operations to make all entries beneath the pivot in the pivot column zero.
    \item Repeat steps 1-3 on the submatrix consisting of all rows below the most recently obtained pivot until we reach the bottom row or there are no remaining nonzero rows.

        NOTE: After step 4, our matrix will be in REF.
    \item Beginning with the rightmost pivot and working left and upwards, use scaling operations to set the pivot to 1.
    \item Use row replacement operations to make all entries above the pivot in the pivot column zero.
    \item Repeat steps 5-6 until all pivots are 1 and there are no nonzero entries in the pivot columns.

        NOTE: Our matrix will be in RREF.
\end{enumerate}

\subsection*{Existence and Uniqueness Theorem}

A linear system is \textit{consistent} \textit{\textbf{if and only if}} the rightmost column of the augmented matrix is not a pivot column.

If a linear system is consistent, then:

\begin{enumerate}
    \item It has a unique solution when there is no free variable.
    \item It has infinitely many solutions when there is a free variable.
\end{enumerate}

% TODO: Rouche-Capelli

\subsection*{Rouche-Capelli Theorem}

A linear system is \textit{consistent} \textit{\textbf{if and only if}}:

\medskip
{\centering

$ \textrm{rank}(A) = \textrm{rank}([ A \hspace{4pt} b ]) $

}
\medskip

If a linear system is consistent, then:

\begin{enumerate}
    \item It has a unique solution when $ \textrm{rank}(A) = n $.
    \item It has a infinitely many solutions when $ \textrm{rank}(A) < n $.
\end{enumerate}

Where $ n $ is the number of columns in the matrix $ A $.

\section*{Lecture 3}

\textbf{Vector}: Ordered list of numbers, or a matrix with only one row or one column.

\subsection*{Vector Operations}

For vectors $ u $, $ v $, $ w $ in $ \mathbb{R}^n $ and for scalars $ c $ and $ d $:

\begin{enumerate}
    \item $ u + v = v + u $
    \item $ (u + v) + w = u + (v + w) $
    \item $ u + 0 = 0 + u = u $
    \item $ u + (-u) = 0 $
    \item $ c(u + v) = cu + cv $
    \item $ (c + d)u = cu + du $
    \item $ (cd)u = c(du) $
    \item $ 1u = u $
\end{enumerate}

\subsection*{Linear Combinations}

Given vectors $ v_1, v_2, \hdots, v_p \in \mathbb{R}^n $ and scalars $ c_1, c_2, ..., c_p $, the vector $ b $ defined by

\medskip
{\centering

$ b = c_1v_1 + c_2v_2 + \hdots + c_pv_p $

}
\medskip

is a \textbf{linear combination} of $ v_1, v_2, \hdots , v_p $ with \textbf{weights} $ c_1, c_2, ... , c_p $.

\bigskip

A solution $ b $ can be generated by a linear combination of columns of $ A $ if and only if there exists a solution to the linear system $ Ax = b $.

% TODO: Examples & Theorems

\subsection*{Spans}

$ \textrm{Span} \{ v_1, \hdots, v_p \} = \{ c_1v_1 + ... + c_pv_p \hspace{4pt} | \hspace{4pt} c_1, ..., c_p \textrm{are scalars.}\} $ is the subset of $ \mathbb{R}^n $ \textbf{spanned} by $ v_1, v_2, ..., v_p $. It is the collection of all linear combinations from $ v_1 $ to $ v_p $.

To ask if a vector is within a span of vectors is the same as asking if the matrix formed from the column vectors has a solution.

% TODO: Examples for Span

\bigskip

Reduce the matrix to reduced echelon form. The rank of the matrix determines the type of the span:
\begin{itemize}
    \item 0 Pivots - Point
    \item 1 Pivot - Line (All points from the origin to the vector)
    \item 2 Pivots - Plane
    \item 3+ Pivots - $ \mathbb{R}^3 $
\end{itemize}

\section*{Lecture 4}

\subsection*{Matrix-Vector Product}

Let $ A = \begin{bmatrix}
    a_1 & a_2 & \hdots & a_n
\end{bmatrix} $ be an $ m \times n $ matrix and $ x \in \mathbb{R}^n $. The \textbf{product} of $ A $ and $ x $ is the linear combination of the columns of $ A $ using the corresponding entries in $ x $ as weights.

\subsection*{Row-Vector Rule}

Each element of the resulting vector is found by taking the \textit{dot product} of a row from the first matrix and a column from the second matrix.

\subsection*{Matrix Equations}

A \textbf{matrix equation} is an equation of the form $ Ax = b $, where $ A $ is a matrix. They can be represented as a linear system, vector equations, or an augmented matrix.

\subsection*{Test for Consistency Theorem}

Let $ A $ be an $ m \times n $ matrix. The following are logically equivalent:

\begin{enumerate}
    \item The matrix equation $ Ax = b $ is consistent.
    \item The vector equation $ x_1a_1 + x_2a_2 + \hdots + x_na_n = b $ has a solution.
    \item $ b $ is in $ \textrm{Span} \{ a_1, a_2, \hdots, a_n \} $.
    \item The linear system with augmented matrix $ \begin{bmatrix}a_1 & a_2 & \hdots & a_n & b \end{bmatrix} $ is consistent.
\end{enumerate}

\bigskip

In summary, the product of two matrices is \textit{consistent} for all $ b $ only if there is a pivot in every row.

\section*{Lecture 5}

\subsection*{Homogeneous Systems}

A \textbf{homogeneous} linear system can be written in the form $ Ax = 0 $, where $ A $ is an $ m \times n $ matrix and $ 0 \in \mathbb{R}^m $ is the zero vector.

\begin{itemize}
    \item Always has at least one solution $ x = 0 $ (\textbf{trivial solution})
    \item Always consistent
    \item \textbf{Nontrivial Solution} $ \leftrightarrow $ 1 or more free variables
\end{itemize}

We can thus derive the following theorems regarding homogeneous systems. Let $ f $ be the number of free variables in the system:

\begin{enumerate}
    \item $ f = 0 \rightarrow $ \textbf{trivial solution} $ x = 0 $ with solution set $ \textrm{Span} \{ 0 \} $
    \item $ f = 1 \rightarrow $ line with solution set $ \textrm{Span} \{ v \} $
    \item $ f = 2 \rightarrow $ plane with solution set $ \textrm{Span} \{ u, v \} $
\end{enumerate}

\subsection*{Non-homogeneous Linear Systems}

A \textbf{non-homogeneous} linear system can be written in the form $ Ax = b $, where $ b \neq 0 $.

\begin{itemize}
    \item Solution sets of a homogeneous system represented by $ Ax = 0 $ shifted by a particular solution $ p $.
    \item The particular solution $ p $ can be found by:
        \begin{enumerate}
            \item Reducing the matrix to RREF.
            \item Setting free variables to $ 0 $.
        \end{enumerate}
\end{itemize}

\section*{Lecture 6}

\subsection*{Linear Independence}

A set of vectors $ \{v_1, v_2, \hdots, v_p \} $ in $ \mathbb{R^n} $ is \textbf{linearly independent} if the vector equation

\medskip
{\centering

$ x_1v_1 + x_2v_2 + \hdots + x_pv_p = 0 $

}
\medskip

has only the trivial solution.
\bigskip

Thus, an $ m \times n $ augmented matrix $ [A \:|\: b] $, where $ b $ is the zero column vector, is linearly independent if:
\begin{itemize}
    \item There are no nontrivial solutions (only the trivial solution).
    \item There are no free variables.
    \item There are pivots equal to $ m $, the number of rows, in the matrix.
\end{itemize}

\subsection*{Special Cases for Linear Dependence}

\begin{enumerate}
    \item A set of one vector: $ \{ v_1 \} $
        \medskip

        $ \{ v_1 \} $ is linearly independent $ \leftrightarrow v_1 \neq 0 $ 
    \item A set of two vectors: $ \{ v_1, v_2 \} $
        \medskip

        $ \{v_1, v_2 \} $ is linearly independent $ \leftrightarrow v_2 \neq cv_1 $
        \medskip

        Thus, a set of two vectors is linearly independent \textbf{iif} the two vectors are not \textit{scalar multiples} of each other.
    \item A set of two or more vectors: $ \{ v_1, v_2, \hdots, v_p \} $
        \medskip

        The set $ S = \{ v_1, v_2, \hdots, v_p \} $ is linearly independent only if \textbf{all} vectors are \textit{not} linear combinations.
        \medskip

        Thus, a set of two or more vectors is linearly independent \textbf{iif} there are no vectors that are \textit{scalar multiples} of each other.
\end{enumerate}

% TODO: a few more special cases must be documented here

\section*{Lecture 7}

This table contains all of the standard matrices for every common matrix transformation.
\begin{table}[h]
    \begin{center}

        \begin{tabular}{|c|c|}
            \hline
            Identity Matrix / No Transformation & $ \begin{bmatrix}
                1 & 0                                               \\
                0 & 1                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Reflection across $ x $-axis & $ \begin{bmatrix}
                1 & 0                                               \\
                0 & -1                                              \\
                                                    \end{bmatrix} $ \\
            \hline
            Reflection across $ y $-axis & $ \begin{bmatrix}
                -1 & 0                                               \\
                0 & 1                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Reflection across $ y = x $ & $ \begin{bmatrix}
                0 & 1                                               \\
                1 & 0                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Reflection across $ y = -x $ & $ \begin{bmatrix}
                0 & -1                                               \\
                -1 & 0                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Reflection across origin & $ \begin{bmatrix}
                -1 & 0                                               \\
                0 & -1                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Horizontal Shear & $ \begin{bmatrix}
                1 & k                                               \\
                0 & 1                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Vertical Shear & $ \begin{bmatrix}
                1 & 0                                               \\
                k & 1                                               \\
                                                    \end{bmatrix} $ \\
            \hline
            Counter-clockwise Rotation & $ \begin{bmatrix}
                \cos(\theta) & -\sin(\theta)               \\
                \sin(\theta) &  \cos(\theta)               \\
                                                    \end{bmatrix} $ \\
            \hline
        \end{tabular}

    \end{center}
\end{table}

\section*{Lecture 8}

\subsection*{Geometric Linear Transformations}

If $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is a linear transformation, we can write $ T(x) = Ax $ , where $ A = \begin{bmatrix} T(e_1) & T(e_2) & \hdots & T(e_n) \end{bmatrix} $ is the $ m \times n $ standard matrix.

\subsection*{Onto and One-to-One Transformations}

A mapping $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is \textbf{onto} if the equation $ T(x) = b $ has a solution for each $ b \in \mathbb{R}^m $. That is, if $ T(x) = b $ can be treated as a \textbf{consistent} linear system for $ b \in \mathbb{R}^m $, it is \textbf{onto}.
\bigskip

A mapping $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is \textbf{one-to-one} if the equation $ T(x) = b $ has a solution for each $ b \in \mathbb{R}^m $ and the solution is \textbf{unique}. That is, $ T(x) = b $ is the image of exactly one $ x \in \mathbb{R}^n $.

\subsection*{Theorem: One-to-One Linear Transformations}

If the mapping $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is a linear transformation, then

\medskip
{\centering

$ T $ is \textbf{one-to-one} $ \leftrightarrow T(x) = 0 $ has only the \textbf{trivial solution}.

}
\medskip

That is, if the linear transformation $ T $ can be written as a homogeneous system $ T(x) = 0 $ with a trivial solution, the solution set is a \textbf{point}, and thus is \textbf{unique}.

\subsection*{Theorem: One-to-One Relation to Matrix Columns}

If the mapping $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is a linear transformation and the matrix $ A $ is the standard matrix of $ T $, then

\begin{itemize}
    \item $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is \textbf{onto} $ \leftrightarrow $ the columns of $ A $ span $ \mathbb{R}^m $.
    \item $ T $ is \textbf{one-to-one} $ \leftrightarrow $ the columns of $ A $ are linearly independent.
\end{itemize}

\section*{Lecture 9}

\subsection*{Matrices}

\underline{Recall}: We have used the following notation for an $ m \times n $ matrix $ A $:

\medskip
{\centering

    $ A = \begin{bmatrix} a_1 & \hdots & a_j & \hdots & a_n \end{bmatrix} $

}
\medskip

where

\medskip
{\centering

$ a_j = \begin{bmatrix} a_{1j} \\ \vdots \\ a_{ij} \\ \vdots \\ a_{mj} \end{bmatrix} \in \mathbb{R}^m $.

}
\medskip

\subsection*{Matrix Equality}

Two matrices are \textbf{equal} if they have the same size and their corresponding entries are equal.

\subsection*{Sums and Scalar Multiples of Matrices}

Let $ A = \begin{bmatrix} a_{ij} \end{bmatrix} $ and $ B = \begin{bmatrix} b_{ij} \end{bmatrix} $ be two $ m \times n $ matrices, and let $ r $ be a scalar. Then:

\begin{itemize}
    \item $ A + B = \begin{bmatrix} a_{ij} + b_{ij} \end{bmatrix} $
    \item $ rA = \begin{bmatrix} ra_{ij} \end{bmatrix} $
\end{itemize}

That is, the \textbf{sum of two matrices} is defined as the $ m \times n $ matrix with entries of sums of all corresponding entries. This property is only defined between two matrices of \textit{the same size}.

\subsection*{Matrix Multiplication}

Let $ A $ be an $ m \times n $ matrix, $ B $ be an $ n \times p $ matrix, and $ x \in \mathbb{R}^p $. The \textbf{composition of linear transformations}

\medskip
{\centering

$ T: \mathbb{R}^p \rightarrow \mathbb{R}^n \rightarrow \mathbb{R}^m $

$ x \rightarrow Bx \rightarrow A(Bx) $

}
\medskip

and be condensed to the linear transformation

\medskip
{\centering

$ T: \mathbb{R}^p \rightarrow \mathbb{R}^m $

$ x \rightarrow A(Bx) $

}
\medskip

Thus, if $ A $ is an $ m \times n $ matrix and $ B $ is an $ n \times p $ matrix, the \textbf{matrix product} $ AB $ consists of concatenated column vectors created from the \textbf{dot product} of every row of $ A $ with every column of $ B $.
\medskip

\textcolor{gray}{Note: Matrix multiplication is only applicable between a matrix $ A $ with $ n $ columns and a matrix $ B $ with $ n $ rows.}

\subsection*{Special Matrices}

The \textbf{main diagonal} of a matrix $ A = \begin{bmatrix} a_{ij} \end{bmatrix} $ is the list of entries $ a_{ii} $.

\begin{itemize}
    \item A matrix $ D_{n \times n} $ is a \textbf{diagonal matrix} if its non-diagonal entries are zero.
    \item A matrix $ L_{n \times n} $ is a \textbf{lower triangular matrix} if its non-zero entries are only in the lower triangle of the matrix.
    \item A matrix $ U_{n \times n} $ is a \textbf{upper triangular matrix} if its non-zero entries are only in the upper triangle of the matrix.
\end{itemize}

\section*{Module 1 Overview}

The following table contains common two-way equivalencies for matrices and linear systems.
\begin{table}[h]
\begin{center}
    \scriptsize

        \begin{tabularx}{\linewidth}{|X|X|X|X|}
           \hline
           $ Ax = b $ & $ A = m \times n $ matrix & $ b = m \times p $ matrix & $ x = n \times p $ matrix \\
           \hline
            \textbf{Consistent} & $ Ax = b $ has a solution & $ b \in \textrm{Span}\{a_1, a_2, \hdots, a_n \} $ & $ \textrm{rank}(A) = \textrm{rank}(\begin{bmatrix} A \:|\: b \end{bmatrix}) $ \\
            \hline
            \textbf{Inconsistent} & $ Ax = b $ has no solutions & $ b \not\in \textrm{Span}\{a_1, a_2, \hdots, a_n \} $ & $ \textrm{rank}(A) \not= \textrm{rank}(\begin{bmatrix} A \:|\: b \end{bmatrix}) $ \\
           \hline
            \textbf{Linear Independence} & $ Ax = 0 $ has \textbf{one solution} & $ T $ is \textbf{one-to-one} & $ 0 \in \textrm{Span}\{a_1, a_2, \hdots, a_n \} $ \\
           \hline
            $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is \textbf{onto} & $ T(x) = b $ \textbf{Consistent} for $ b \in \mathbb{R}^m $ & columns of $ A $ span $ \mathbb{R}^m $ & range $ \subseteq $ codomain  \\
            \hline
            $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ is \textbf{one-to-one} & $ T(x) = b $ \textbf{one solution} for $ b \in \mathbb{R}^m $ & $ T(x) = b $ \textbf{linearly independent} for $ b \in \mathbb{R}^m $ & range $ = $ codomain \\
            \hline
            $ T: \mathbb{R}^n \rightarrow \mathbb{R}^m $ & $ n $ columns = domain & $ m $ rows = codomain & $ T $ represented as $ m \times n $ matrix \\
            \hline
        \end{tabularx}

\end{center}
\end{table}
\FloatBarrier

\begin{itemize}
    \item (\textbf{Homogeneous System} $ \leftrightarrow $ $ Ax = 0 $ $) \rightarrow $ \textbf{Consistent}
\end{itemize}

\part*{Unit 2}

\section*{Lecture 10}

\subsection*{Tranpose}

The tranpose of $ A $ (where $ A $ is an $ m \times n $ matrix), denoted $ A^T $, is the $ m \times n $ matrix where the columns of $ A $ become the rows of $ A^T $.

\medskip
{\centering

$ A = \begin{bmatrix}
    3 & 1 & 0 \\
    2 & 1 & 3 \\
\end{bmatrix} $,
$ A^T = \begin{bmatrix}
    3 & 2 \\
    1 & 1 \\
    0 & 3 \\
\end{bmatrix} $

}
\medskip

\subsubsection*{Properties of Tranposes}

\begin{enumerate}
    \item $ (A^T)^T = A $
    \item $ (A + B)^T = A^T + B^T $
    \item $ (rA)^T = rA^T $ for any scalar $ r $
    \item $ (AB)^T = B^TA^T $
\end{enumerate}

\subsection*{Invertible}

An $ m \times n $ matrix $ A $ is \textbf{invertible} if there is an $ m \times n $ matrix $ B $ such that:


\medskip
{\centering

    $ AB = I_n = BA $

}
\medskip

where $ I_n $ is the $ n \times n $ identity matrix.
\medskip

Otherwise, if $ A $ is \textit{not} invertible, it is \textbf{singular}.

\subsubsection*{Inverse}

The \textbf{inverse} of $ A $, denoted by $ A^{-1} $, is the unique matrix such that:

\medskip
{\centering

    $ AA^{-1} = I_n = A^{-1}A $

}
\medskip

Given
$ A = \begin{bmatrix}
    a & b \\
    c & d \\ 
\end{bmatrix} $.
The \textbf{inverse} of $ A $ is defined as
$ \displaystyle A^{-1} = \frac{1}{ad - bc}
    \begin{bmatrix}
    d & -b \\
    -c & a \\
    \end{bmatrix} $.
\bigskip

\subsubsection*{Theorem: Invertibility}

$ A $ is \textbf{invertible} if and only if the augmented matrix
$ \begin{bmatrix}
A & I \\
\end{bmatrix} $ is \textbf{consistent}.

Thus, given a matrix $ A $, you can solve the augmented matrix $ \begin{bmatrix} A & I \end{bmatrix} $. The solution matrix $ x $ is $ A^{-1} $. 

\subsubsection*{Invertible: Example 1}

Given $ A = \begin{bmatrix}
    -2 & 3 \\
    3 & -5 \\
\end{bmatrix} $ and $ B = \begin{bmatrix}
    -5 & -3 \\
    -3 & -2 \\
\end{bmatrix} $, show that $ B = A^{-1} $.
\bigskip

To show that $ A $ and $ B $ are inverses of each other, their matrix product must be the $ 2 \times 2 $ identity matrix.
\bigskip

\medskip
{\centering

$ \begin{bmatrix}
    -2 & 3 \\
    3 & -5 \\
\end{bmatrix} \begin{bmatrix} 
    -5 & -3 \\
    -3 & -2 \\
\end{bmatrix} = \begin{bmatrix}
    1 & 0 \\
    0 & 1 \\
\end{bmatrix} $

}
\medskip

\subsubsection*{Invertible: Example 2}

Let $ A = \begin{bmatrix}
    1 & -3 \\
    -2 & 10 \\
\end{bmatrix} $
\begin{enumerate}
    \item Find the inverse of $ A $.
        \bigskip

        The easiest way is to use the formula $\displaystyle \frac{1}{ad - bc} \begin{bmatrix}
            d & -b \\
            -c & a \\
        \end{bmatrix} $.
        \bigskip

        The inverse of $ A $ is $\displaystyle \frac{1}{10 - 6} \begin{bmatrix}
            10 & 2 \\
            3 & 1 \\
        \end{bmatrix} = \begin{bmatrix}
            \frac{10}{4} & \frac{2}{4} \\
            \frac{3}{4} & \frac{1}{4} \\
        \end{bmatrix} $
    \item Solve the system $ Ax = b $, where $ b = \begin{bmatrix}
            -4 \\ 8 \\
    \end{bmatrix} $
    \bigskip

    Just solve the augmented matrix.
        $ \begin{amatrix}{2}
            1 & 0 & -4 \\
            0 & 1 & 0 \\
        \end{amatrix} $
\end{enumerate}

\subsubsection*{Invertible: Example 3}

Find the inverse of the matrix $ A $ if it exists.

\begin{enumerate}
    \item $ A = \begin{bmatrix}
            1 & 0 & -2 \\
            0 & 1 & 4 \\
            0 & -2 & -8 \\
    \end{bmatrix} $
        \bigskip

        Solve the augmented matrix $ \begin{amatrix}{1}
            A & I \\
        \end{amatrix} $.
        \bigskip

        \medskip
        {\centering

        $ \begin{bmatrix}
            1 & 0 & -2 & 1 & 0 & 0 \\
            0 & 1 & 4 & 0 & 1 & 0 \\
            0 & -2 & -8 & 0 & 0 & 1 \\
        \end{bmatrix} $
        \bigskip

        $ \begin{bmatrix}
            1 & 0 & -2 & 1 & 0 & 0 \\
            0 & 1 & 4 & 0 & 0 & -\frac{1}{2} \\
            0 & 0 & 0 & 0 & 1 & \frac{1}{2} \\
        \end{bmatrix} $
        \bigskip


        }
        \medskip

        Since the augmented matrix is \textbf{inconsistent}, the matrix $ A $ is \textbf{not invertible}.

    \item $ A = \begin{bmatrix}
            1 & 0 & -2 \\
            0 & 1 & 4 \\
            0 & -2 & -7 \\
    \end{bmatrix} $
        \bigskip

        \medskip
        {\centering

        $ \begin{bmatrix}
            1 & 0 & -2 & 1 & 0 & 0 \\
            0 & 1 & 4 & 0 & 1 & 0 \\
            0 & -2 & -7 & 0 & 0 & 1 \\
        \end{bmatrix} $
        \bigskip

        $ \begin{bmatrix}
            1 & 0 & 0 & 1 & 4 & 2 \\
            0 & 1 & 0 & 0 & -7 & -4 \\
            0 & 0 & 1 & 0 & 2 & 1 \\
        \end{bmatrix} $

        }
        \medskip

        Since the augmented matrix is \textbf{consistent}, the inverse of matrix $ A $ is $ \begin{bmatrix}
            1 & 4 & 2 \\
            0 & -7 & -4 \\
            0 & 2 & 1 \\
        \end{bmatrix} $.
\end{enumerate}

\section*{Lecture 11}

\subsection*{Determinant}

\textit{Recall}: For a $ 1 \times 1 $ matrix, the \textbf{determinant} is defined as $ \textrm{det} \: A = \: \textrm{det} \: a_{11} $.
\bigskip

\textit{Recall}: For a $ 2 \times 2 $ matrix $ A =
\begin{bmatrix}
    a & b \\
    c & d \\
\end{bmatrix} $, the \textbf{determinant} of the matrix is $ \textrm{det} \: A = ad - bc $.
\bigskip

The \textbf{determinant} for an $ n \times n $ matrix, where $ n \geq 2 $, is the sums of $ n $ terms of the form $ \pm a_{1j} \: \textrm{det} \: A_{1j} $, where the entries $ a_{11}, a_{12}, \hdots, a_{1n} $ are from the first row of $ A $.
\medskip

The submatrices in the form $ A_{ij} $ are obtained from the matrix $ A $ by deleting the $ i $-th row and the $ j $-th column.

\medskip
{\centering

$ A = \begin{bmatrix}
    1 & 5 & 0 \\
    2 & 4 & -1 \\
    0 & -2 & 0 \\
\end{bmatrix} $
\medskip

$ \textrm{det} \: A = a_{11} \: \textrm{det} \: A_{11} - a_{12} \: \textrm{det} \: A_{12} + a_{13} \: \textrm{det} \: A_{13} $
\medskip

$ \textrm{det} \: A = a_{11}
\: \textrm{det} \begin{bmatrix}
    a_{22} & a_{23} \\
    a_{32} & a_{33} \\
\end{bmatrix}
- a_{12}
\: \textrm{det} \begin{bmatrix}
    a_{21} & a_{23} \\
    a_{31} & a_{33} \\
\end{bmatrix}
+ a_{13}
\: \textrm{det} \begin{bmatrix}
    a_{21} & a_{22} \\
    a_{31} & a_{32} \\
\end{bmatrix} $

\medskip


$ \textrm{det} \: A = 1(0 - 2) - 5(0 - 0) + 0(-4 - 0) = -2 $

}
\medskip

\subsubsection*{Cofactors}

The \textbf{$ (i, j) $-cofactor} of matrix $ A $ is the number $ C_{ij} $ given by:

\medskip
{\centering

$ C_{ij} = (-1)^{i + j} \: \textrm{det} \: A_{ij} $

}
\medskip

\subsubsection*{Cofactor Expansion}

Notice that the cofactors along the first row can be used to define the \textbf{determinant}:

\medskip
{\centering

$ \textrm{det} \: A = a_{11} C_{11} + a_{12}C_{12} + \hdots + a_{1n}C_{1n} $

}
\medskip

\subsubsection*{Theorem: Cofactor Expansion}

The determinant of an $ n \times n $ matrix $ A $ can be computed by a \textbf{cofactor expansion} across \textit{any row or column}.

\medskip
{\centering

    $ \textrm{det} \: A = a_{i1}C_{i1} + a_{i2}C_{i2} + \hdots + a_{in}C_{in} $ (expansion across row)
    \medskip

    $ \textrm{det} \: A = a_{1j}C_{1j} + a_{2j}C_{2j} + \hdots + a_{nj}C_{nj} $ (expansion across column)

}
\medskip

\subsubsection*{Determinants: Example 1}

Compute the determinant for the matrix $ \begin{bmatrix}
    0 & 0 & 0 \\
    1 & 2 & 3 \\
    4 & 5 & 6 \\
\end{bmatrix} $.
\bigskip

By using \textbf{cofactor expansion} across the first row, the elements in the first row become the scalars to the determinants of the submatrices. The determinant will be 0.

\subsubsection*{Determinants: Example 2}

Compute the determinant for the matrix $ \begin{bmatrix}
    1 & 0 & 0 & 0 \\
    1 & -2 & 0 & 0 \\
    1 & -2 & 3 & 0 \\
    1 & -2 & 3 & -4 \\
\end{bmatrix} $
\bigskip
% TODO: finish this example sometime

% TODO: the determinant of the matrix is equal to the determinant of its transpose (proof it)

\section*{Lecture 12}

% TODO: Cramer's rule

\section*{Lecture 13}

\subsection*{Vector Spaces}

Let $ V $ be a nonempty set of vectors on which two operations are defined:

\medskip
{\centering

$ u + v \in V $ (closed under addition)
\medskip

$ cu \in V $ (closed under scalar multiplication)

}
\medskip

The set $ V $ with the operations form a \textbf{vector space} if the following eight axioms hold for all $ u $, $ v $, $ w $ in $ V $ and for all scalars $ c $ and $ d $:

\begin{enumerate}
    \item $ u + v = v + u $ 
    \item $ (u + v) + w = u + (v + w) $
    \item $ u + \vec{0} = u $
    \item For each $ u \in V $, there is a vector $ -u $ such that $ u + (-u) = 0 $.
    \item c(u + v) = cu + cv
    \item (c + d)u = cu + du
    \item c(du) = (cd)u
    \item 1u = u
\end{enumerate}

Using the axioms listed above, we can show that:

\begin{enumerate}
    \item The zero vector is unique.
    \item The negative of $ u $, denoted by $ -u $, is unique.
\end{enumerate}

\subsubsection*{Vector Spaces: Examples}

\begin{enumerate}
    \item $ \mathbb{R}^n = \begin{Bmatrix}
            \begin{bmatrix}
                x_1 \\
                x_2 \\
                \vdots \\
                x_n
            \end{bmatrix} & : & x_i \in \mathbb{R}
    \end{Bmatrix} $
\item $ \mathbb{M}_{x \times n} = \begin{Bmatrix}
        [a_{ij}] \: | \:  a_{ij} \in \mathbb{R}
\end{Bmatrix} $
    \item $ \mathbb{P}_n = \textrm{the set of polynomials of degree at most} n \textrm{with real coefficients} $
\end{enumerate}

\subsection*{Subspaces}

Let $ H $ be a subset of a vector space $ V $. $ H $ is called a \textbf{subspace} of $ V $ if it satisfies the following conditions:

\begin{enumerate}
    \item The zero vector of $ V $ is in $ H $.
    \item $ H $ is closed under vector addition:
        For any $ u $, $ v \in H \rightarrow (u + v) \in H $
    \item $ H $ is closed under scalar multiplication:
        For any $ u \in H $, $ c \in \mathbb{R} \rightarrow cu \in H $
\end{enumerate}

Any subspace of a subspace of $ V $ is a subspace of $ V $.
\bigskip

\textbf{WARNING}: $ \mathbb{R}^2 $ is not a subspace of $ \mathbb{R}^3 $ .

\subsubsection*{Zero Subspace}

The \textbf{zero subspace} $ \begin{Bmatrix} 0 \end{Bmatrix} $ is the set consisting of only the zero vector of a vector space $ V $.
    
\subsection*{Theorem: Subspace Spanned by Vectors}
If $ v_1, \hdots, v_p $ are in a vector space $ V $, then $ H = \textrm{Span} \begin{Bmatrix}
    v_1, \hdots, v_p \\
\end{Bmatrix} $ is a subspace of $ V $. Thus, \textbf{spans are subspaces}.

\subsection*{Spanning Sets}

Let $ V $ be a vector space and $ v_i \in V $. Then, let $ S = \begin{Bmatrix}
    v_1, \hdots, v_p
\end{Bmatrix} $ be a set of vectors, and $ H = \textrm{Span} \begin{Bmatrix}
    v_1, \hdots, v_p
\end{Bmatrix} $.

\begin{itemize}
    \item The set $ S $ is a \textbf{spanning set} for $ H $.
    \item $ H $ is called the \textbf{subspace of $ V $ spanned} by the set $ S $.
\end{itemize}

\section*{Lecture 14}

\subsection*{Null Space}

The \textbf{null space} of $ m \times n $ matrix $ A $, denoted by $ \textrm{Nul} \: A $, is the solution set of $ Ax = 0 $.

\section*{Lecture 15}

\subsection*{Column Space}

The \textbf{column space} of an $ m \times n $ matrix $ A $, denoted by $ \textrm{Col} \: A $, is the set of all linear combinations of $ A $.

\subsubsection*{Theorem: Column Space is the Span}

The \textbf{column space} of a $ m \times n $ matrix $ A $ is equivalent to the \textbf{span} of the columns of $ A $.

\subsubsection*{Theorem: Column Space}

Given an $ m \times n $ matrix $ A $, a $ m \times 1 $ matrix $ w $ is \textbf{in} the \textbf{column space} of $ A $ if $ Ax = w $ is a consistent system.

\subsection*{Theorem: Column Space of Transpose}

Given an $ m \times n $ matrix $ A $, the \textbf{row space} of $ A^T $ is the same as the \textbf{column space} of $ A $, because the rows of $ A^T $ are the columns of $ (A^T)^T = A $.

\subsubsection*{Theorem: Vector Subspaces}
Given an $ m \times n $ matrix $ A $, then:

\begin{enumerate}
    \item $ \textrm{Nul} (A) $ is a subspace of $ \mathbb{R}^n $
    \item $ \textrm{Col} (A) $ is a subspace of $ \mathbb{R}^m $
\end{enumerate}

\subsection*{Theorem: Geometric Description of Nul A, Row A, and Col A}

\begin{itemize}
    \item $ \textrm{Nul} \: A \perp \textrm{Row} \: A $
    \item $ \textrm{Nul} \: A^T \perp \textrm{Col} \: A $
\end{itemize}

\section*{Lecture 16}

\subsection*{Bases}

Let $ H $ be a subspace of a vector space $ V $ and let $ b_i \in V $. $ B = \begin{Bmatrix}
    b_1, \hdots, b_p \\
\end{Bmatrix} $ is a \textbf{basis} for $ H $ if:

\begin{enumerate}
    \item $ B $ is \textbf{linearly independent}.
    \item $ H = \textrm{Span} \begin{Bmatrix}
            b_1, \hdots, b_p \\
    \end{Bmatrix} $ 
\end{enumerate}

\subsection*{Theorem: Basis from Invertible Matrices}

The columns of an $ n \times n $ invertible matrix $ A $ form a basis for $ \mathbb{R}^n $.

\subsection*{Theorem: Basis is Spanning Set}

Assume that $ B $ is the \textbf{basis} of matrix $ H $. Thus, $ B $ is also the spanning set of $ H $.

\subsection*{Spanning Set Theorem}

Let $ S = \begin{Bmatrix}
    v_1, \hdots, v_p \\
\end{Bmatrix} $ be a set in $ V $ and let $ H = \textrm{Span} \begin{Bmatrix}
    v_1 \hdots, v_p \\
\end{Bmatrix} $

\begin{enumerate}
    \item If one of the vectors in $ S $ is a linear combination of the other vectors in $ S $, then $ S $ still spans $ H $ if we remove that vector from $ S $.
    \item If one of the vectors in $ S $ is \textbf{linearly dependent} on the other vectors in $ S $, then that vector can be removed from $ S $ and $ S $ will still span $ H $.
    \item If $ H \neq \begin{Bmatrix}
            0
    \end{Bmatrix} $, some subset of $ S $ is a basis for $ H $.
\end{enumerate}

\subsubsection*{Spanning Set Theorem: Example 1}

Consider if $ H = \textrm{Span} \begin{Bmatrix}
    1, t, t^2, 2 - 3t^2 \\
\end{Bmatrix} $, then $ H = \textrm{Span} \begin{Bmatrix}
    1, t, t^2
\end{Bmatrix} $.
\medskip

A \textbf{basis} can be formed from a \textbf{spanning set} by discarding unneeded vectors.

\subsubsection*{Spanning Set Theorem: Example 2}

\medskip
{\centering

Let $ v_1 = \begin{bmatrix}
    1 \\
    0 \\
    0 \\
\end{bmatrix} $,
$ v_2 = \begin{bmatrix}
    0 \\
    1 \\
    1 \\
\end{bmatrix} $,
$ v_3 = \begin{bmatrix}
    -3 \\
    4 \\
    4 \\
\end{bmatrix} $.

}
\medskip

\begin{enumerate}
    \item Determine if the set $ \begin{Bmatrix}
            v_1, v_2, v_3 \\
    \end{Bmatrix} $ is a basis for $ \mathbb{R}^3 $.
    \bigskip

        The set $ \begin{Bmatrix}
            v_1, v_2, v_3 \\
        \end{Bmatrix} $ is \textbf{linearly dependent}, and thus does not row-reduce into an $ n \times n $ identity matrix. Therefore, the set is not a basis for $ \mathbb{R}^3 $.
    \item Let $ H = \textrm{Span} \begin{Bmatrix}
            v_1, v_2, v_3 \\
    \end{Bmatrix} $ be a subspace for $ \mathbb{R}^3 $. Then, $ H = \textrm{Span} \begin{Bmatrix}
        v_1, v_2, \\
    \end{Bmatrix} $ and $ \begin{Bmatrix}
        v_1, v_2 \\
    \end{Bmatrix} $ is a basis for $ H $.
    \bigskip

        Since $ v_3 $ is a \textbf{linearly dependent} vector of $ v_1 $ and $ v_2 $, by the \textbf{Spanning Set Theorem} we can remove this vector from the set and it will still span $ H $. Thus, it will still be a \textbf{basis} for $ H $.
\end{enumerate} 

\subsection*{Basis for Null Space}

Find the \textbf{basis} for $ \textrm{Nul} \: A \neq 0 $, we solve for $ Ax = 0 $ by using the row reduction algorithm.

\subsection*{Basis for Row Space}

Reduce matrix $ A $ to RREF, then the \textbf{nonzero rows} are linearly independent and form a basis for $ \textrm{Row} \: A $.

\subsection*{Basis for Column Space}

The \textbf{pivot columns} of matrix $ A $ form a basis for $ \textrm{Col} \: A $.

\subsubsection*{Theorem: Basis for Column Space}

\begin{enumerate}
    \item The pivot columns of $ A $ are \textbf{linearly independent}.
    \item Each non-pivot column of $ A $ is a linear combination of the preceding pivot columns.
\end{enumerate}

\section*{Lecture 17}

\subsection*{Dimension}

The \textbf{dimension}, denoted by $ \textrm{dim} \: V $, is the number of vectors in a basis for a nonzero vector space $ V $.

\begin{itemize}
    \item The \textbf{dimension} of the zero vector is \textbf{0}.
    \item If $ V $ is spanned by a \textbf{finite} set, then $ V $ is \textbf{finite-dimensional}.
    \item If $ V $ cannot be spanned finite set, then $ V $ is \textbf{infinite-dimensional}.
\end{itemize}

\subsubsection*{Dimension: Examples}

\begin{enumerate}
    \item $ V = \mathbb{R}^n $
        \bigskip

        $ E = \begin{Bmatrix}
            e_1 & e_2 & e_3 & \hdots & e_n \\
        \end{Bmatrix} $
        \bigskip

        Thus, $ \textrm{dim} \: V = n $.
    \item $ V = \mathbb{P}_3 $
        \bigskip

        $ \textrm{dim} \: V = 4 $
    \item $ V = \mathbb{P} $
        \bigskip

        The vector space $ V $ would be \textbf{infinite-dimensional}.
    \item $ V = \mathbb{M}_{2 \times 2} $
        \bigskip

        $ \textrm{dim} \: V = m \times n = 4 $
\end{enumerate}

\subsection*{Theorem: Dimension to Rank}

The \textbf{dimension} of a vector space $ V $ is equivalent to the rank of $ V $.

\subsection*{Theorem: Linearly Dependent Spaces}

If a vector space $ V $ has a basis $ B = \begin{Bmatrix}
    b_1, \hdots, b_n \\
\end{Bmatrix} $, then any set in $ V $ containing more than $ n $ vectors must be \textbf{linearly dependent}.
\medskip

This also follows from the \textbf{Spanning Set Theorem}; any unnecessary vectors will be \textbf{linearly dependent}.

\subsection*{Theorem: Dimensions for Bases}

If both $ B = \begin{Bmatrix}
    b_1, \hdots, b_n \\
\end{Bmatrix} $ and
$ B' = \begin{Bmatrix}
    b'_1, \hdots, b'_m \\
\end{Bmatrix} $ are bases for a vector space $ V $, then $ n = m $.

\subsection*{Theorem: Subspace of Finite-Dimensional Space}

Let $ H $ be a subspace of a finite-dimensional space $ V $. Any linearly independent set in $ H $ can be expanded to a basis for $ H $. $ H $ is also finite-dimensional and $ \textrm{dim} \: H \leq \textrm{dim} \: V $.

\section*{Lecture 18}

\subsection*{Unique Representation Theorem}

Given a basis $ B = \begin{Bmatrix}
    b_1, \hdots, b_n \\
\end{Bmatrix} $ for a vector space $ V $, then every vector $ x \in V $ can be expressed as a linear combination $ x = c_1b_1 + \hdots + c_nb_n $ in exactly one way.

\section*{Lecture 19}

\subsection*{Eigenvector}

An \textbf{eigenvector} of an $ n \times n $ matrix $ A $ is a nonzero vector $ x $ such that $ Ax = \lambda x $ or some scalar $ \lambda $.
\bigskip

To determine whether $ v $ is an \textbf{eigenvector} of $ A $:

\begin{enumerate}
    \item Compute $ Av $.
    \item If $ Av $ is a scalar multiple of $ v $, then $ v $ is an \textbf{eigenvector} of $ A $.
    \item The \textbf{eigenvalue} is the scalar factor.
\end{enumerate}

\bigskip

\textit{Example.} Let $ A = \begin{bmatrix}
    2 & 3 \\
    -1 & 6 \end {bmatrix} $, determine whether $ \vec{x} = \begin{bmatrix}
    3 \\ 1 \\
    \end{bmatrix} $ is an \textbf{eigenvector} of $ A $.

\begin{enumerate}
    \item $ A \vec{x} = \begin{bmatrix}
        2 & 3 \\
        -1 & 6 \\
    \end{bmatrix} \begin{bmatrix}
        3 \\ 1 \\
    \end{bmatrix} = \begin{bmatrix}
        6 + 3 \\
        -3 + 6 \\
    \end{bmatrix} = \begin{bmatrix}
        9 \\ 3 \\
    \end{bmatrix} $
    \item You can see that the matrix $ A\vec{x} = \begin{bmatrix}
        9 \\ 3 \\
\end{bmatrix} $ is an \textbf{eigenvector} of $ A $ with an \textbf{eigenvalue} of $ \lambda = 3 $.
\end{enumerate}

\bigskip

\textit{Example.} Let $ A = \begin{bmatrix}
    2 & 3 \\
    -1 & 6 \\
\end{bmatrix} $. Show that $ \lambda = 3 $ is an \textbf{eigenvalue} of matrix $ A $ and find the corresponding \textbf{eigenvectors}.

\begin{enumerate}
    \item $ A \vec{x} = \lambda \vec{x} $
        \medskip

        $ A \vec{x} - \lambda \vec{x} = 0 $
        \medskip

        $ ( A - \lambda I ) \vec{x} = 0 $
        \medskip
    \item $ (\begin{bmatrix} 
            2 & 3 \\
            -1 & 6 \\
    \end{bmatrix} - \begin{bmatrix}
        3 & 0 \\
        0 & 3 \\
    \end{bmatrix}) \vec{x} = 0 $
\item $ \begin{bmatrix}
        -1 & 3 \\
        -1 & 3 \\
\end{bmatrix} \vec{x} = 0 $
\item Solve the homogenous system by reducing to row echelon form.
    $ \begin{amatrix}{2}
        1 & -3 & 0 \\
        0 & 0 & 0
    \end{amatrix} $
    \medskip

    $ \vec{x} = \begin{bmatrix}
        3 x_2 \\
        x_2 \\
    \end{bmatrix} = x_2 \begin{bmatrix}
        3 \\
        1 \\
    \end{bmatrix} $
    \medskip
\item The vector $ \begin{bmatrix}
        3 \\
        1 \\
\end{bmatrix} $ is therefore also a \textbf{basis} of the \textit{eigenspace}.

\end{enumerate}

\section*{Lecture 21}

\subsection*{The Diagonalization Theorem}

Given a matrix $ A = PDP^{-1} $, with diagonal matrix $ D $,\textit{if and only if} the columns of $ P $ are $ n $ linearly independent eigenvectors of $ A $.
\medskip

Furthermore, the diagonal entries of $ D $ are the eigenvalues of $ A $ that correspond to the eigenvectors in $ P $.


\end{document}
