\documentclass[12pt, letterpaper]{article}

\usepackage{../spreamble}

\usepackage{tikz}
\usetikzlibrary{trees}


\title{COP3530}
\author{Ian Zou}

\begin{document}

\maketitle

\part{Algorithm Analysis}

\section*{What is an Algorithm?}

An \textbf{algorithm} is a step-by-step procedure for solving a problem.

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Algorithm}          & \textbf{Program}                                  \\ \hline
Design-focused              & Implementation-focused                            \\ \hline
Implementation-independent  & The implementation                                \\ \hline
Theory                      & Practice                                          \\ \hline
\end{tabular}
\end{center}

\section*{Benchmarking}

Not all algorithms perform equally; there are some algorithms that are much more efficient than other \textit{naive} algorithms.

\section*{Measuring Performance}

We can measure an algorithm's performance in a plethora of ways:

\begin{itemize}
    \item Time
    \item Space
    \item Energy Consumption
    \item Error Rates (Approximation Algorithms)
\end{itemize}

\subsection*{Approach 1 - Timing}

Simply mark the time of the algorithm's start and end, and take the \textit{delta time} to find out how long the algorithm took. For example, in C you could use the \texttt{timespec} struct, or in C++ you could use \texttt{std::clock}.

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Pros}     & \textbf{Cons}                                     \\ \hline
Easy to measure   & Machine-dependent                                 \\ \hline
Easy to interpret & Compiler-dependent                                \\ \hline
                  & Implementation-dependent                          \\ \hline
                  & Unpredictable for small inputs                    \\ \hline
                  & Input dependence not captured                     \\ \hline
\end{tabular}
\end{center}

\subsection*{Approach 2 - Counting}

Count the number of times each operation will be performed by interpretting the code.

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Pros}               & \textbf{Cons}                         \\ \hline
Machine-independent         & All operations are equal              \\ \hline
Captures input dependence   & Tedious to compute                    \\ \hline
                            & Implementation-dependent              \\ \hline
                            & Time not captured                     \\ \hline
\end{tabular}
\end{center}

\subsection*{Approach 3 - Asymptotic Behavior / Order of Growth}

We cannot always measure algorithms in absolute units (such as seconds for time). Different hardware may result in different times taken for an algorithm to complete, and to compare two results with extraneous factors is meaningless.

Instead, we measure in terms of their \textit{growth rate}; as the load increases, how much mores resources will we need to complete the algorithm?

\subsubsection*{Time Complexity}

\textbf{Time complexity} describes the asymptotic growth rate of the algorithm's time required as the input increases.

\subsubsection*{Big-O Notation}
\begin{itemize}
    \item \textbf{Big-O}: Upper Bound
        \begin{itemize}
            \item If there exists two positive constants $ n_0$ and $ c $ such that $ T(n) \leq c \cdot f(n) $ for all $ n \geq n_0 $.
            \item $ f(n) $ is an upper bound on the performance (growth rate of $ T(n) $).
        \end{itemize}
    \item \textbf{Big-$\Omega$}: Lower Bound
        \begin{itemize}
            \item If there exists two positive constants $ n_0$ and $ c $ such that $ T(n) \geq c \cdot g(n) $ for all $ n \geq n_0 $.
            \item $ g(n) $ is a lower bound on the growth rate of $ T(n) $.
        \end{itemize}
    \item \textbf{Big-$\Theta$}: Upper + Lower Bound
        \begin{itemize}
            \item If $ T(n) = O(g(n)) $ and $ T(n) = \Omega(g(n)) $
            \item $ g(n) $ is a tight upper and lower bound on the growth rate of $ T(n) $.
        \end{itemize}
\end{itemize}

\subsubsection*{Common Growth Rates}
\begin{itemize}
    \item \textbf{Constant Growth Rate} $ O(1) $
        \begin{itemize}
            \item Processing time is \textit{independent} of input $ n $.
        \end{itemize}
    \item \textbf{Linear Growth Rate} $ O(n) $
        \begin{itemize}
            \item Processing time scales linearly with input $ n $.
        \end{itemize}
    \item \textbf{Quadratic Growth Rate} $ O(n^2) $
        \begin{itemize}
            \item Processing time incrases in proportion to the square of input $ n $.
        \end{itemize}
    \item \textbf{Logarithmic Growth Rate} $ O(\log(n)) $
        \begin{itemize}
            \item Processing time incrases in proportion to the $ \log(n) $.
        \end{itemize}
\end{itemize}

\part{Arrays, Lists, \& Queues}

\section*{Abstract Data Types}

\textbf{Abstract Data Type}: Class of objects whose logical behavior is defined by a set of values and a set of operations. Implementation-agnostic.

\section*{Lists}

\begin{itemize}
    \item Ordered collection of data
    \item Elements have some position
    \item Linear Structure
    \item Size, can grow or shrink
\end{itemize}

\subsection*{Arrays}

\begin{itemize}
    \item Fixed size
    \item Contiguous in memory
    \item Stores similar elements
    \item Random access
    \item Constant access time
    \item Expensive to push/pop
\end{itemize}

\subsection*{Linked Lists}

\begin{itemize}
    \item Consists of nodes with pointers to the next element
    \item Not contiguous in memory
    \item Fast to push/pop
    \item No random access
    \item Pointer overhead on traversal
\end{itemize}

\subsection*{ADT: Stacks}

The Stack ADT is characterized by the \textbf{LIFO} principle. Elements are stacked like plates, with the most recent one on the top.

\subsubsection*{Operations}

\begin{itemize}
    \item \texttt{push(item)}: insert an element to top of stack
    \item \texttt{pop()}: removes last inserted element (top of stack)
    \item \texttt{peek()}: returns last inserted element (C++ STL: \texttt{top()})
    \item \texttt{size()}: returns the number of elements stored
    \item \texttt{empty()}: returns true if the stack size is 0
\end{itemize}

\subsubsection*{Stacks: Array Implementation}

\begin{itemize}
    \item Constant-time operations
    \item Fixed size, no random access
\end{itemize}

\subsubsection*{Stacks: Linked List Implementation}

\begin{itemize}
    \item Constant-time operations
    \item Variable size
    \item Requires more memory
    \item More overhead (due to indirection with pointers to heap)
\end{itemize}

\section*{ADT: Queues}

The Queue ADT is characterized by the \textbf{FIFO} principle. Analogous to a line, the first person to enter the queue is the first person to be removed.

\subsubsection*{Operations}

\begin{itemize}
    \item \texttt{enqueue(item)}: inserts an element to the back of queue
    \item \texttt{dequeue()}: removes first inserted element (front of queue)
    \item \texttt{size()}: returns the number of elements stored
    \item \texttt{empty()}: returns true if the queue size is 0
\end{itemize}

\part{Trees}

Trees are rooted, directed, acyclic structures.

\section*{Properties of Trees}

\begin{itemize}
    \item One root
    \item Each node has one parent
    \item No cycles
\end{itemize}

\section*{Terminology of Trees}

\begin{itemize}
    \item \textbf{Parent}: Predecessors of a node.
    \item \textbf{Children}: Successors of a node.
    \item \textbf{Root}: Top node of the tree.
        \begin{itemize}
            \item Root nodes have \textit{no parents}.
        \end{itemize}
    \item \textbf{Leaf}: Bottom nodes of the tree.
        \begin{itemize}
            \item Leaf nodes have \textit{no children}.
        \end{itemize}
    \item \textbf{Level (Depth)}: Distance of a node from the root.
        \begin{itemize}
            \item The root node is \textit{Level 0}.
        \end{itemize}
    \item \textbf{Height}: Number of nodes in the longest path from the root to a leaf.
\end{itemize}

\section*{N-Ary Trees}
\begin{itemize}
    \item All properties of a \textbf{tree}, plus:
    \item Each node consists of at most \textbf{n} children.
\end{itemize}

\section*{Binary Tree}
\begin{itemize}
    \item All properties of a \textbf{tree}, plus:
    \item Each node consists of at most \textbf{2} children.
\end{itemize}

\subsection*{Binary Tree Types}
\begin{itemize}
    \item \textbf{Perfect Binary Tree}
        \begin{itemize}
            \item Full binary tree of height \textbf{$ h $} (0-based).
            \item Exactly $ 2^{h+1} - 1 $ nodes.
            \item A binary tree that is complete and full is perfect.
        \end{itemize}
    \item \textbf{Complete Binary Tree}
        \begin{itemize}
            \item Perfect binary tree through \textbf{$ h - 1 $}.
            \item Extra leaf nodes at level \textbf{$ h $}, towards the left.
        \end{itemize}
    \item \textbf{Full Binary Tree}
        \begin{itemize}
            \item Each node must either have 2 nodes or 0 nodes.
        \end{itemize}
\end{itemize}

\section*{Balanced Trees}

To reduce the occurrence of worst case tree patterns that degrade a BST's time complexity to a worst case of \texttt{O(n)}, we can modify the tree when we encounter some of these patterns to optimize the tree.

\subsection*{AVL Trees}

\begin{itemize}
    \item \textbf{Balance Factor}: left subtree height - right subtree height
    \item In a balanced tree, every node has a \textbf{balance factor} property of \texttt{0}, \texttt{-1}, or \texttt{1}.
\end{itemize}

\subsubsection*{Rotations}

To keep an AVL Tree balanced, rotations are made when we encounter certain patterns that degrade the performance of the tree by increasing a node's balance factor greater than \texttt{1} or less than \texttt{-1}.

\begin{itemize}
    \item \textbf{Left-Left Case}:
        \begin{itemize}
            \item Parent: \texttt{+2}
            \item Child: \texttt{+1}
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{3}
                    child { node {2}
                        child { node {1}
                        }
                        child [missing]
                    }
                    child [missing]
                    ;
                \end{tikzpicture}

                \medskip
            \item \textbf{Fix}: Perform a \textbf{right rotation}.
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{2}
                    child { node {1}
                    }
                    child { node { 3 }
                    }
                    ;
                \end{tikzpicture}
        \end{itemize}
    \item \textbf{Right-Right Case}:
        \begin{itemize}
            \item Parent: \texttt{-2}
            \item Child: \texttt{-1}
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{1}
                    child [missing]
                    child { node {2}
                        child [missing]
                        child { node {3}
                        }
                    }
                    ;
                \end{tikzpicture}

                \medskip
            \item \textbf{Fix}: Perform a \textbf{left rotation}.
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{2}
                    child { node {1}
                    }
                    child { node { 3 }
                    }
                    ;
                \end{tikzpicture}
        \end{itemize}
    \item \textbf{Left-Right Case}:
        \begin{itemize}
            \item Parent: \texttt{+2}
            \item Child: \texttt{-1}
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{3}
                    child { node {1}
                    child [missing]
                        child { node {2}
                        }
                    }
                    child [missing]
                    ;
                \end{tikzpicture}

                \medskip
            \item Fix: Perform a \textbf{left} rotation on the child, followed by a \textbf{right} rotation on the parent.

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{3}
                    child { node {2}
                        child { node {1}
                        }
                        child [missing]
                    }
                    child [missing]
                    ;
                \end{tikzpicture}
                \medskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{2}
                    child {
                        node {1}
                    }
                    child {
                        node {3}
                    };
                \end{tikzpicture}
        \end{itemize}
    \item \textbf{Right-Left Case}:
        \begin{itemize}
            \item Parent: \texttt{-2}
            \item Child: \texttt{+1}
                \bigskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{1}
                    child [missing]
                    child { node {3}
                        child { node {2}
                        }
                        child [missing]
                    }
                    ;
                \end{tikzpicture}

                \medskip
            \item Fix: Perform a \textbf{right} rotation on the child, followed by a \textbf{left} rotation on the parent.

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{1}
                    child [missing]
                    child { node {2}
                        child [missing]
                        child { node {3}
                        }
                    };
                \end{tikzpicture}
                \medskip

                \begin{tikzpicture}[nodes={draw, circle}, ->]
                    \node{2}
                    child {
                        node {1}
                    }
                    child {
                        node {3}
                    };
                \end{tikzpicture}
        \end{itemize}
\end{itemize}

\bigskip

When do we need to rotate nodes in an AVL tree?

\begin{enumerate}
    \item 
\end{enumerate}

\subsection*{Red-Black Trees}

\begin{itemize}
    \item A node can be \textbf{red} or \textbf{black}.
    \item Root node is \textbf{black}.
    \item \textbf{Red} nodes always has \textbf{black} children.
    \item \textbf{Null} nodes are defined as \textbf{black}.
    \item Much cheaper than height comparison due to color flips.
\end{itemize}

\subsubsection*{Red-Black Tree: Insertion}

\begin{enumerate}
    \item Insert the node into BST as a \textbf{red} node.
    \item If:
        \begin{itemize}
            \item Tree is \textbf{empty}: insert as \textbf{black}
            \item \textit{Parent} is \textbf{black}: do nothing
            \item \textit{Parent} is \textbf{red}:
                \begin{itemize}
                    \item \textit{Uncle} is \textbf{red}: invert colors of parent, uncle, and grandparent
                    \item \textit{Uncle} is \textbf{black}: rotate (similar to AVL Trees) % TODO: link it up
                        % TODO: add examples for invariance correction
                \end{itemize}
        \end{itemize}
    \item Cascade rule up the parent chain to check and remediate invariance of the tree.
\end{enumerate}

\subsubsection*{Red-Black Tree: Use Case}

Red-black trees are used to back \textit{hash maps} when collisions occur. When poor hash indices are generated and cause collision, using a List ADT to handle collisions will degrade the time complexity to \texttt{O(n)}. By using a red-black tree, we can mitigate the worst-case time complexity to \texttt{O(log(n))}.

\subsection*{Splay Tree}

Self-adjusting binary search tree with additional property that recent accessed elements are quick to access.

\begin{itemize}
    \item \textbf{Splaying}: series of rotations that push a node to the root of the tree
    \item \textit{Splaying} is done after a node is accessed or inserted
    \item Guarantees \texttt{m} tree operations will take \texttt{O(m log(n))} time.
    \item On average, a tree operation is \texttt{O(log(n))}.
    \item In the worst case, the splay tree is skewed left or right, and thus tree operations are \texttt{O(n)}. Splaying will drastically speed up subsequent operations.
\end{itemize}

\subsubsection*{Splay Tree: Use Case}

Due to the Splay Tree's self-balancing nature and ability to make recently accessed items extremely fast to access, it is implemented primarily in \textbf{cache} and \textbf{garbage collection}.

\subsection*{B Tree}

% TODO

\section*{Heaps}

\textbf{Heaps} are \textit{complete binary tree} data structures that satisfy the \textbf{heap property}.

\begin{itemize}
    \item \textbf{Heap Property}: Values of children must be greater than or equal to its own value.
    \item \textbf{Min Heap}: Value of the root node is the smallest.
    \item \textbf{Max Heap}: Value of the root node is the greatest.
\end{itemize}

\subsection*{Heaps: Use Case}

Heaps are usually used to implement \textbf{priority queues}, where the maximum or minimum is always the root node of the tree.
\medskip

Heaps can also be useful when implementing \textbf{Djikstra's Algorithm}.

\subsection*{Heaps: Operations}

\begin{itemize}
    \item \texttt{heapify}: The heapify operation is used to heal the heap and maintain the \textit{heap property}. Parent nodes are compared to their children, and are swapped if they do not adhere to the \textit{heap property}, and this is done to the entire tree.
    \item \texttt{insert}: Use \textit{level order traversal} to find the next empty node in the current level. If the tree is \textit{full}, then the current level will not contain an empty node, in which case we move down a level.
        \medskip

        Insert the value in the next empty node, then \textit{heapify} down the path of the new node.
        \bigskip

        \textbf{Time Complexity}: \texttt{O(log(n))} down one path.

\end{itemize}


\end{document}

